Linear Algebra as a Unifying Theory\\of Wireless Workload Modeling

### 2008-02-05 paper.tex
Graphs!!!

%\begin{abstract}
%\input{abstract.tex}
%\end{abstract}

%\section{Algebraic Representations \\of Network Behavior}
\section{The General Matrix Model}
\label{sec:representations}

@@

\subsection{Network: Marginal Behavior}
\label{sec:marginal-behavior}

\noindent
...

@@

\subsection{Uniformity}

@@

The matrix model has the benefit of making the uniformization of behaviors completely explicit. It also allows us to easily define intermediate choices. If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
\begin{align}
F' = ((1-\alpha)(SS^t)+\alpha(DD^t))F.
\end{align}
The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors. Source-only behavior is given by $\alpha=0$, for destination-only set $\alpha=1$, and for the standard average, set $\alpha=\frac{1}{2}$.

@@

We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in local area networks (\caps{LAN}s). Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematic precision and succinctness, while simultaneously being made clearer and more general.

Most of what are called traffic models are actually only partial descriptions of one or a handful of aspects of a complete model. The ubiquitous constant bit-rate (\caps{CBR}) model, for example, merely describes the nature of the packets in each flow in a workload: in a given flow, the sizes of packets and the intervals between them are are all uniform. Hence, the transmitted bit-rate is constant.

Two things are require for the direct empirical comparison of traffic models:
\begin{itemize}
\item An objective standard for comparing quality of models.
% WinMee/Broadnets provide this
\item A common framework in which different models can be implemented and compared.
% this paper provides this
\end{itemize}

Despite great breakthroughs in understanding and methodology, the networking community remains without a cookbook for generating realistic \caps{LAN} workloads. Worse still, it is unclear what ``realistic'' should mean in this context. Foremost, a clear and testable definition of realism for traffic models is needed. In the absence of such a definition, it remains standard in wireless experiments to generate traffic using the simplest possible model: all nodes, flows and packets are assumed to behave uniformly. Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily. While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better. It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.

To address this foundational problem in network traffic modeling, we have introduced a fundamentally different approach to evaluating the realism of workload models~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics. We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values similar to those observed using the original trace to generate traffic. Using this new approach, we explore the space of traffic models, establishing rigorously that the commonly used uniform models for node, flow and packet behavior drastically misrepresent important \caps{WLAN} performance metrics at all levels of the protocol stack.

It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}. Here, however, we face another fundamental challenge in traffic modeling. Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads. The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two end-points, however, it remains unclear how to assign end-points to flows such that the marginal distribution of flow per node matches the BiPareto model indicated. Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes. We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid. Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.

Our point is not to denigrate the models proposed in~\cite{Hernandez06:wlan-traffic} in any way. On the contrary: these parametric models are the state of the art in modeling traffic patterns for wireless local area networks. The purpose, rather, is to point out how large a gap still remains between having accurate high-level models of marginal aspects of network behavior and being able to generate usable workloads with those models. Moreover, a great deal of ambiguity lies in that gap since there is no single obvious way to combine models.

To clarify, disambiguate and standardize the specification of generative local area traffic models, we propose the \textit{Traffic Object Model} (\textit{\caps{TOM}}). The \caps{TOM} entails a hierarchy of abstract classes: \class{Node}, \class{Flow}, and \class{Packet} are the primary classes; the \class{Network} class collects everything. Each abstract class defines an interface of properties. A traffic model defines concrete classes at each level that provide the required interfaces; each class defines its own properties by delegation, inheritance, or overriding properties of other classes. Because of the separation of the \caps{TOM} into well-defined levels, each concrete class naturally defines a sub-model, specific to that level's behavior; sub-models can be mixed and matched combinatorially.

We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network. The standard na\"ive model uses the \class{Uniform} sub-model at all levels. The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels. The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.

The rest of the paper is organized as follows. In \Section{motivation} we present motivation and related work. Our experimental and analytical methodology is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.

This work extends that published in~\cite{Karpinski07:realism}~and~\cite{Karpinski07:cbr-failure}, using the same methodology for defining and quantifying the realism of synthetic traffic models. The Traffic Object Model builds on the combinatorial approach taken in~\cite{Karpinski07:cbr-failure}, while providing an flexible and general object-oriented framework for constructing generative traffic models. The traffic models evaluated here overlap and augment those evaluated in both prior works; comparison with the models proposed in {\FHC} is entirely new. The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, client-server, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows. We have also implemented performance metric aggregation based on the \caps{IETF} framework for \caps{IP} performance metrics~\cite{rfc:ip-metrics,rfc:jitter}.

\section{Methodology}\label{sec:methodology}

The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results. Clearly, we need not simulate subatomic particle interactions in a wireless network simulator. Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same. Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior. This requirement leads us to the following definition:

@@

This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required. We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.

@@

Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of synthetic traffic models defined using the Traffic Object Model (\caps{TOM}) framework. We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004. The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes. The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed. We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th. This trace contains a broad variety of behaviors and entails a very large volume of traffic: 2.1 million flows, 58 million packets, and 52 billion bytes.
% 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.

@@

\begin{table}[t]
\vspace{0.75em}
\begin{center}
\begin{tabular}[t]{|l|}
\hline
\begin{minipage}[l]{2.7in}
\begin{small}
\vspace{0.7em}
\input{properties.tex}
%\vspace{0.6em}
\end{small}
\end{minipage}
\\\hline
\end{tabular}
\end{center}
\vspace{-0.5em}
\caption{{\footnotesize{TOM}} base classes and their attributes.}
\label{tab:properties}
\vspace{-2em}
\end{table}

\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:tom}

\begin{table*}[t]
\begin{center}
\small
\input{models.tex}
\footnotesize\\[0.5em]
%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
\end{center}
\vspace{-0.75em}
\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
\label{tab:models}
\vspace{-1em}
\end{table*}

In this section, we present a generic object-oriented framework for implementing traffic generation models. This framework is called the Traffic Object Model (\caps{TOM}). It defines four levels of responsibility: network, node, flow and packet. The framework is designed so that different behaviors at each level can be naturally implemented  as classes in object-oriented programming languages. The framework automatically gains the advantages of object orientation, in particular modularity and encapsulation. Every traffic model is naturally split into separate spheres of responsibility, which constitute sub-models for the aspects of the model's behavior at each layer. Encapsulation and polymorphism of objects, allows us to mix-and-match the sub-models from different models to create new hybrid behaviors. It is trivial, for example, to produce trace-based node behavior, coupled with parametric flow behavior, and uniform packet behavior. Not all combinations of sub-models are valid, however; which classes may be combined depends on the details of each class.

% TODO: fix-up client-server vs. source-destination terminology.

Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be served either by an actual abstract class or by an interface. In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if abstract classes are desired, they can be simulated by creating base classes with stub methods that throw informative exceptions upon being called.} 
Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work. Future traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper. Concrete subclasses implement the behavior of specific traffic models at each level. A complete synthetic workload is instantiated as follows:
\begin{enumerate}
\item create a \class{Network} object of the appropriate subclass;
\item the top-level \class{Network} object in turn creates a collection of \class{Node} objects and associated \class{Flow} objects;
\item each \class{Flow} object generates a stream of \class{Packet} objects.
\end{enumerate}
The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation. Our implementation does the latter, generating a collection of Qualnet configuration files and trace files specific to each workload instantiation (see \Section{simulations}).

\subsection{Traffic Models}\label{sec:models}

For each of the \class{Node}, \class{Flow} and \class{Packet} abstract superclasses, we define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The \class{Trace} classes take properties of objects directly from the behavior of the corresponding object in the original trace. The \class{Nonparametric} and \class{Parametric} classes use models of the corresponding type to randomly generate the properties of objects. The \class{Uniform} classes generate identical objects with average properties. The specific details of how each the concrete classes at each level are defined are specified in Table~\ref{tab:models}. The parameters for the fitted models used in the \class{Parametric} subclasses are estimated from the corresponding nonparametric distribution using maximum likelihood estimation. Figure~\ref{fig:trace-vs-uniform} shows visual depictions comparing trace versus uniform behavior at various levels in example scenarios.

While the modularity of the Traffic Object Model allows us to mix and match sub-models, not all possible combinations are valid. In particular, note that the \class{Trace} classes are tied to a specific object from the underlying trace, whereas for other classes, there is no such correspondence to any specific object. Consequently, once a non-trace model has been used, trace models may not be used at the lower levels. This still leaves 40 valid composite combinations of sub-models; we only evaluate twelve of these. Composite models are abbreviated by the initials of their constituent sub-models. The fully trace-base model, \class{TTT}, serves as an experimental control. We consider \class{TTN}, \class{TTP}, and \class{TTU} in order to compare models of packet behavior. We evaluate different flow models, using the best possible model at the packet level: \class{TNN}, \class{TPN}, \class{TUN}. Similarly, we compare models for node behavior: \class{NNN}, \class{PNN}, \class{UNN}. Finally, we use the na\"ive uniform model, \class{UUU}, and the composite parametric model, \class{PPP}, proposed by {\FHC}~\cite{Hernandez06:wlan-traffic}, as external points of comparison.

@@

The 24-hour trace is split into 144 10-minute segments, each of which serves as the basis for a set of simulations using different traffic models. The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model. Various partially synthetic intermediate models, described in \Section{models}, are simulated to study the impact of different aspects of traffic behavior on network performance. To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models. The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model. Moreover, the following features are preserved across all models: the numbers of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).

@@

\begin{figure*}
\begin{centering}
\subfloat[App. Average Jitter: {\footnotesize{STU}}]
{\includegraphics[width=1.75in]{application-average-jitter-stu}%
\label{fig:application-average-jitter-stu}}
\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
\label{fig:network-packet-drop-rate-utu}}
\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
\label{fig:network-packet-drop-rate-stu}}
\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
\label{fig:mac-retransmission-failures-suu}}
\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}). Each data point represents a pair of matched simulations: the $x$-coordinate is the metric value for trace traffic, the $y$-coordinate is the metric value for the synthetic model. Scenarios are plotted according to the amount of flows in their trace segment: the bottom third with filled dots; the middle third with circles; the top third with squares.}
\label{fig:scatterplots}
\end{centering}
\vspace{-1em}
\end{figure*}

>>>>

\begin{abstract}
We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks. Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general. We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally. We conclude with future directions for modeling research based on the new algebraic framework.
\end{abstract}

%Most of what are called traffic models are only partial descriptions of one or a handful of aspects of a total pattern of traffic in a network. The ubiquitous constant bit-rate (\caps{CBR}) model, for example, merely describes the nature of the packets in each flow in a workload: in a given flow, the sizes of packets and the intervals between them are are all uniform. Hence, the transmitted bit-rate is constant.

%Two things are require for the direct empirical comparison of traffic models:
%\begin{itemize}
%\item An objective standard for comparing quality of models.
%% WinMee/Broadnets provide this
%\item A common framework in which different models can be implemented and compared.
%% this paper provides this
%\end{itemize}

Despite great breakthroughs in understanding and methodology, the networking community remains without a cookbook for generating realistic \caps{LAN} workloads. Drastic assumptions of uniformity abound in typical experimental workload generation. Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily. While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better. It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.

We believe that two major breakthroughs are needed to bring a much needed revolution to the world of traffic modeling. The first requirement is an objective standard of realism by which to compare traffic models. Such a standard has already been introduced in~\cite{Karpinski07:realism} and used to demonstrate that uniform traffic modeling, including the ubiquitous \caps{CBR} model, drastically distort performance metrics at all levels of the network protocol stack~\cite{Karpinski07:cbr-failure}. The second required breakthrough is a unified framework for expressing traffic models that allows researchers to systematically express and understand the assumptions and transformations entailed in each model.

In this paper, we present a highly general model for representing detailed full-network traffic patterns algebraically. The General Matrix Model (\caps{GMM}) represents the behavior or each network flow as a vector in a high-dimensional linear space, and represents total network behavior as a large matrix whose rows are the behavior vectors. The crucial property of this representation is that it respects vector space operations: summation of behavior vectors gives an accurate description of the aggregate behavior of the flows represented by those vectors. The benefit of this representation is that it makes it immediately possible to express a broad variety of assumptions and conditions on traffic patterns in terms of simple matrix operations. It also allows provides great mathematical insight and allows us to bring all the tools of modern linear algebra to bear on the problem of traffic modeling.

%Worse still, it is unclear what ``realistic'' should mean in this context. Foremost, a clear and testable definition of realism for traffic models is needed. In the absence of such a definition, it remains standard in wireless experiments to generate traffic using the simplest possible model: all nodes, flows and packets are assumed to behave uniformly. Flows are commonly modeled as constant bit-rate (\caps{CBR}) streams of identical packets sent at an invariant rate; the flows in an experiment are commonly given the same duration and data rate; nodes in the network are often selected as flow endpoints randomly or arbitrarily. While more complex approaches have been proposed~\cite{Sommers04,Avallone04,Avallone06}, % TODO: better references here.
%the simplest, most na\"ive models remain standard because no objective criterion exists by which to compellingly argue that other models are any better. It is simply unclear that simplistic models do any harm, or how, if at all, complex models provide benefit.

%To address this foundational problem in network traffic modeling, we have introduced a fundamentally different approach to evaluating the realism of workload models~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. Rather than subjectively choosing statistical measures that may or may not actually influence network performance, the realism of a model is measured directly in terms of its ability to accurately reproduce critical performance metrics. We define a traffic model to be \textit{sufficiently realistic} with respect to a given performance metric, if the model produces metric values similar to those observed using the original trace to generate traffic. Using this new approach, we explore the space of traffic models, establishing rigorously that the commonly used uniform models for node, flow and packet behavior drastically misrepresent important \caps{WLAN} performance metrics at all levels of the protocol stack.

%It is natural to compare these common, na\"ive models with the high-level parametric models proposed by \FHC~\cite{Hernandez06:wlan-traffic}. Here, however, we face another fundamental challenge in traffic modeling. Even with specific models for various aspects of network behavior, it remains non-trivial to combine them into a coherent composite model capable of generating concrete workloads. The {\FHC} models, for example, specify the distribution of flows per node in the network; since every flow has two end-points, however, it remains unclear how to assign end-points to flows such that the marginal distribution of flow per node matches the BiPareto model indicated. Similarly, the models dictate BiPareto marginal distribution of flow sizes across the entire network, but say nothing about how these sizes are to be distributed among the nodes. We could assume that the number of flows per node is independent from the size of the flows at that node, but, as we shall see, this assumption is invalid. Moreover, the details of flow behavior---packet sizes and inter-packet intervals---remain entirely undefined.

%Our point is not to denigrate the models proposed in~\cite{Hernandez06:wlan-traffic} in any way. On the contrary: these parametric models are the state of the art in modeling traffic patterns for wireless local area networks. The purpose, rather, is to point out how large a gap still remains between having accurate high-level models of marginal aspects of network behavior and being able to generate usable workloads with those models. Moreover, a great deal of ambiguity lies in that gap since there is no single obvious way to combine models.

%To clarify, disambiguate and standardize the specification of generative local area traffic models, we propose the \textit{Traffic Object Model} (\textit{\caps{TOM}}). The \caps{TOM} entails a hierarchy of abstract classes: \class{Node}, \class{Flow}, and \class{Packet} are the primary classes; the \class{Network} class collects everything. Each abstract class defines an interface of properties. A traffic model defines concrete classes at each level that provide the required interfaces; each class defines its own properties by delegation, inheritance, or overriding properties of other classes. Because of the separation of the \caps{TOM} into well-defined levels, each concrete class naturally defines a sub-model, specific to that level's behavior; sub-models can be mixed and matched combinatorially.

%We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network. The standard na\"ive model uses the \class{Uniform} sub-model at all levels. The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels. The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.

The rest of the paper is organized as follows. In \Section{motivation} we present motivation and related work. The General Matrix Model is presented in \Section{general-matrix-model}. Our experimental and analytical methodology for evaluating traffic models presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.

%This work extends that published in~\cite{Karpinski07:realism}~and~\cite{Karpinski07:cbr-failure}, using the same methodology for defining and quantifying the realism of synthetic traffic models. The Traffic Object Model builds on the combinatorial approach taken in~\cite{Karpinski07:cbr-failure}, while providing an flexible and general object-oriented framework for constructing generative traffic models. The traffic models evaluated here overlap and augment those evaluated in both prior works; comparison with the models proposed in {\FHC} is entirely new. The Qualnet simulation methodology has been enhanced by the implementation of full-duplex, client-server, trace-driven raw \caps{IP}, \caps{TCP}, and \caps{UDP} flows. We have also implemented performance metric aggregation based on the \caps{IETF} framework for \caps{IP} performance metrics~\cite{rfc:ip-metrics,rfc:jitter}.

%\section{Algebraic Representations \\of Network Behavior}
\section{The General Matrix Model}
\label{sec:general-matrix-model}
\label{sec:representations}

@@

%\subsection{Network: Marginal Behavior}
%\label{sec:marginal-behavior}

%\noindent
%...

@@

\subsection{Uniform Models}

@@

The matrix model has the benefit of making the uniformization of behaviors completely explicit. It also allows us to easily define intermediate choices. If we wish make the behavior of each flow a weighted average of the aggregate behaviors of its source and destination nodes, we can express this easily:
\begin{align}
F' = \frac{1}{f}((1-\alpha)(SS^t)+\alpha(DD^t))F.
\end{align}
The parameter $\alpha$ allows us to smoothly vary the weighting of the source and destination behaviors. Source-only behavior is given by $\alpha=0$; destination-only by $\alpha=1$; to weight source and destination equally, take $\alpha=\frac{1}{2}$.

@@

\section{Experimental Methodology}\label{sec:methodology}

To evaluate whether traffic models are realistic or not, we use the differential simulation methodology introduced in~\cite{Karpinski07:realism} and used in~\cite{Karpinski07:cbr-failure} to compare a variety of uniform and trace-based models. This approach is based on the following definition, initially proposed in~\cite{Karpinski07:realism} and formalized in \cite{Karpinski07:cbr-failure}:

@@

The essential idea behind the differential simulation technique is to compare traffic produced by synthetic models with known examples of actual real-world trace traffic in a variety of experimental scenarios. To accomplish this, we begin with a large 24-hour example of wireless trace traffic, break it into 144 ten-minute scenarios, and use each scenario's behavior as a control for a set synthetic traces, generated by various synthetic traffic models whose realism we with to evaluate. For each scenario, the synthetic models are configured to approximate the original trace behavior as closely as the models allow. Important network-wide performance metrics are then collected from simulations, and synthetic performance results compared with control results to determine the realism of the synthetic models.

%The art of simulation lies in knowing which details must be realistic and which may be abstracted into simpler, approximate models without affecting the accuracy of the results. Clearly, we need not simulate subatomic particle interactions in a wireless network simulator. Instead we use high-level physical models that approximate the real physics well enough that the performance results are the same. Similarly, when modeling network usage, we must require that our models produce the same results as actual user behavior. This requirement leads us to the following definition:

%This definition depends on many factors: the type of wireless scenario, the performance metrics under consideration, the actual usage behavior used for comparison, and how strong a notion of statistical equivalence is required. We discuss different measures of error and how to evaluate statistical equivalence in Sections~\ref{sec:error-measures}~and~\ref{sec:statistical-equivalence}.

\subsection{Traffic Models}
\label{sec:traffic-models}

The original trace traffic constitutes the first model we simulate, although it is not really a model, {per se}. This serves as a control group against which to compare all other models. Before we can evaluate the General Matrix Model (\caps{GMM}) presented in \Section{general-matrix-model}, we must establish that one potentially important aspect of traffic behavior that is \textit{not} represented in the \caps{GMM} can legitimately be discarded. That aspect of behavior is \textit{\caps{TCP} feedback}. The \caps{GMM} does not specify the \caps{IP} protocol type of each flow. Therefore we must establish the extent to which using \caps{UDP} flows, without reliable delivery or congestion control, affects performance metrics.
%\caps{GMM} also does not indicate which flows are paired with each other in a client-server relationship. While this pairing does not affect non-\caps{TCP} flows, we implement.

@@

Our general methodology is to compare performance metrics in simulations using real traffic patterns from traces to the same metrics in simulations using a variety of trace-based models and synthetic traffic models defined using the General Matrix Model. We use a 24-hour trace recorded in an infrastructured 802.11g wireless \caps{LAN} with 18 access points, deployed at the 60th Internet Engineering Task Force meeting (\caps{IETF}60), held in San Diego during August of 2004. The traffic trace was captured using \texttt{\small{tcpdump}} at a single router, through which all wireless traffic for the meeting was routed, including traffic between wireless nodes. The snap length of the capture was 100 bytes, allowing \caps{IP}, \caps{ICMP}, \caps{UDP} and \caps{TCP} headers to be analyzed. We limit our work to the 24-hour sub-trace recorded on Wednesday, August 4th. This trace contains a broad variety of behaviors and entails a very large volume of traffic: 2.1 million flows, 58 million packets, and 52 billion bytes.
% 2,085,563 flows, 58,377,994 packets of application data, and 51,988,245,537 bytes of application data.

@@

%\begin{table}[t]
%\vspace{0.75em}
%\begin{center}
%\begin{tabular}[t]{|l|}
%\hline
%\begin{minipage}[l]{2.7in}
%\begin{small}
%\vspace{0.7em}
%\input{properties.tex}
%%\vspace{0.6em}
%\end{small}
%\end{minipage}
%\\\hline
%\end{tabular}
%\end{center}
%\vspace{-0.5em}
%\caption{{\footnotesize{TOM}} base classes and their attributes.}
%\label{tab:properties}
%\vspace{-2em}
%\end{table}

%\subsection{The Traffic Object Model~(\caps{TOM})}\label{sec:tom}

%\begin{table*}[t]
%\begin{center}
%\small
%\input{models.tex}
%\footnotesize\\[0.5em]
%%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
%\end{center}
%\vspace{-0.75em}
%\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
%\label{tab:models}
%\vspace{-1em}
%\end{table*}

%In this section, we present a generic object-oriented framework for implementing traffic generation models. This framework is called the Traffic Object Model (\caps{TOM}). It defines four levels of responsibility: network, node, flow and packet. The framework is designed so that different behaviors at each level can be naturally implemented  as classes in object-oriented programming languages. The framework automatically gains the advantages of object orientation, in particular modularity and encapsulation. Every traffic model is naturally split into separate spheres of responsibility, which constitute sub-models for the aspects of the model's behavior at each layer. Encapsulation and polymorphism of objects, allows us to mix-and-match the sub-models from different models to create new hybrid behaviors. It is trivial, for example, to produce trace-based node behavior, coupled with parametric flow behavior, and uniform packet behavior. Not all combinations of sub-models are valid, however; which classes may be combined depends on the details of each class.

%% TODO: fix-up client-server vs. source-destination terminology.

%Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be served either by an actual abstract class or by an interface. In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if abstract classes are desired, they can be simulated by creating base classes with stub methods that throw informative exceptions upon being called.} 
%Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work. Future traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper. Concrete subclasses implement the behavior of specific traffic models at each level. A complete synthetic workload is instantiated as follows:
%\begin{enumerate}
%\item create a \class{Network} object of the appropriate subclass;
%\item the top-level \class{Network} object in turn creates a collection of \class{Node} objects and associated \class{Flow} objects;
%\item each \class{Flow} object generates a stream of \class{Packet} objects.
%\end{enumerate}
%The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation. Our implementation does the latter, generating a collection of Qualnet configuration files and trace files specific to each workload instantiation (see \Section{simulations}).

%\subsection{Traffic Models}\label{sec:models}

%For each of the \class{Node}, \class{Flow} and \class{Packet} abstract superclasses, we define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The \class{Trace} classes take properties of objects directly from the behavior of the corresponding object in the original trace. The \class{Nonparametric} and \class{Parametric} classes use models of the corresponding type to randomly generate the properties of objects. The \class{Uniform} classes generate identical objects with average properties. The specific details of how each the concrete classes at each level are defined are specified in Table~\ref{tab:models}. The parameters for the fitted models used in the \class{Parametric} subclasses are estimated from the corresponding nonparametric distribution using maximum likelihood estimation. Figure~\ref{fig:trace-vs-uniform} shows visual depictions comparing trace versus uniform behavior at various levels in example scenarios.

%While the modularity of the Traffic Object Model allows us to mix and match sub-models, not all possible combinations are valid. In particular, note that the \class{Trace} classes are tied to a specific object from the underlying trace, whereas for other classes, there is no such correspondence to any specific object. Consequently, once a non-trace model has been used, trace models may not be used at the lower levels. This still leaves 40 valid composite combinations of sub-models; we only evaluate twelve of these. Composite models are abbreviated by the initials of their constituent sub-models. The fully trace-base model, \class{TTT}, serves as an experimental control. We consider \class{TTN}, \class{TTP}, and \class{TTU} in order to compare models of packet behavior. We evaluate different flow models, using the best possible model at the packet level: \class{TNN}, \class{TPN}, \class{TUN}. Similarly, we compare models for node behavior: \class{NNN}, \class{PNN}, \class{UNN}. Finally, we use the na\"ive uniform model, \class{UUU}, and the composite parametric model, \class{PPP}, proposed by {\FHC}~\cite{Hernandez06:wlan-traffic}, as external points of comparison.

@@

The 24-hour trace is split into 144 10-minute segments, each of which serves as the basis for a set of simulations using different traffic models. The traffic models range from a completely realistic trace-driven model, to a standard \caps{CBR} traffic model. Various partially synthetic intermediate models, described in \Section{traffic-models}, are simulated to study the impact of different aspects of traffic behavior on network performance. To preserve the fairness of the performance comparison, we keep as many features as possible constant across different traffic models. The traffic generated by each synthetic model preserves as many characteristics from the original trace as possible, within the constraints of the model. Moreover, the following features are preserved across all models: the numbers of wireless nodes, the number of flows, the number of application-initiated data units sent, the total bytes of application data sent, and the average flow duration (and therefore the average data rate).

@@

%\begin{figure*}
%\begin{centering}
%\subfloat[App. Average Jitter: {\footnotesize{STU}}]
%{\includegraphics[width=1.75in]{application-average-jitter-stu}%
%\label{fig:application-average-jitter-stu}}
%\subfloat[Net. Packet Drop Rate: {\footnotesize{UTU}}]
%{\includegraphics[width=1.75in]{network-packet-drop-rate-utu}%
%\label{fig:network-packet-drop-rate-utu}}
%\subfloat[Net. Packet Drop Rate: {\footnotesize{STU}}]
%{\includegraphics[width=1.75in]{network-packet-drop-rate-stu}%
%\label{fig:network-packet-drop-rate-stu}}
%\subfloat[Link Retrans. Failures: {\footnotesize{SUU}}]
%{\includegraphics[width=1.75in]{mac-retransmission-failures-suu}%
%\label{fig:mac-retransmission-failures-suu}}
%\caption{Scatter plots of various metrics for an alternate traffic model ({\footnotesize{STU}}, {\footnotesize{UTU}}, {\footnotesize{SUU}}) versus real trace traffic~({\footnotesize{TTT}}). Each data point represents a pair of matched simulations: the $x$-coordinate is the metric value for trace traffic, the $y$-coordinate is the metric value for the synthetic model. Scenarios are plotted according to the amount of flows in their trace segment: the bottom third with filled dots; the middle third with circles; the top third with squares.}
%\label{fig:scatterplots}
%\end{centering}
%\vspace{-1em}
%\end{figure*}
